Linear Meld~(LM) is a linear logic programming language
aimed for the parallel implementation of graph-based
algorithms~\cite{cruz-iclp14}. Like Datalog, LM is
a \emph{forward-chaining} logic programming language since computation is driven by a set of
inference rules that are used to update a database of logical facts. In Datalog,
programs are monotonic and therefore the database grows in size as more facts are
inferred from the logical rules. In LM, logical
facts are linear and thus can be retracted when a rule is inferred. The use of linear
facts greatly increases the power of the language but also increase the
complexity of its implementation since database facts are removed often.

LM is a high-level declarative language that offers a concise and
expressive framework to define graph based algorithms that are provably correct.
LM has been applied to a wide range of problems and machine learning
algorithms, including: belief propagation~\cite{Gonzalez+al:aistats09paraml},
belief propagation with residual splash~\cite{Gonzalez+al:aistats09paraml},
PageRank, graph coloring, N-Queens, shortest path, diameter estimation,
map reduce, quick-sort, neural network training, minimax, and many others.

In previous work~\cite{cruz-ppdp14}, we have described the implementation of the
virtual machine, including its data structures and how programs are
paralellized. In this paper, we describe our compilation strategy for
transforming logical rules into C++ code. While the initial compiler for the virtual
machine used an identical compilation scheme (not fully described
in~\cite{cruz-ppdp14}), we have refitted the runtime system to allow
stand-alone compilation of programs. Our goal was to reduce the overhead of executing
interpreted byte code and better understand the effectiveness and limitations of the
compilation scheme. Our results show that compiled programs are only 2 to 6
times slower than hand-written C programs.

We present an algorithm that compiles logical rules with comprehensions and
aggregates into efficient C++ code. The compiler supports rule priorities, allowing
the programmer to order rules based on their priority of inference. To the best
of our knowledge, this is the first available compilation strategy for a linear logic
language that supports these 3 features combined. The
contributions of this paper are then three-fold: (1) a novel algorithm to compile
prioritized linear logic rules with aggregates and comprehensions; (2) the
interplay between the database layout and compiled code; and (3) comparison and
analsys by comparing our compilation with
hand-written C programs and interpreted code.

The remainder of the paper is organized as follows. First, we briefly introduce
the LM language. Next, we present an overview of the runtime support available
to compiled rules and then we discuss our contributions which include the
algorithm for compiling rules, comprehensions and aggregates into efficient
iterator-based C code. We then present experimental results comparing our
compiled programs with our old implementation. The paper finishes with related
work and some conclusions.
